On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-24 paper/application.tex
revised application

However, here $\theta$ is an effective or generalized
irradiation level. 
%no longer a direct measure of
%irradiation. 
By Eq. \ref{real_theta}, $\theta$ is now related to the true
irradation temperature $\tirr$ through the function $g$ given by 
Eq. \ref{g_def}. This is
important because a physically non-irradiated disk with $\tirr=0$ will
still have a non-zero effective irradiation, $\theta\neq0$. 
%On the
%other hand, fully irradiated disks with  $T=T_\mathrm{irr}$ have 
%$\theta=1$, as is in the previous problem formulation
%(Eq. \ref{tirr_def}).

@@

\subsection{Inviscid stability condition}
We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
From the results in \S\ref{2d_inviscid}, we conclude that this
system cannot be unstable anywhere without
viscous effects acting on the perturbations if 
\begin{align}
  \gamma > \frac{3}{2} \quad \text{and} \quad Q >
  \sqrt{\frac{6}{5}} \label{ppd_invisc_cond} 
\end{align} 
are both satisfied.

PPDs become optically-thin ($\tau\to 0$) with 
increasing distance from the star. Then $\theta$ decreases, and
the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
5/6$ in the outer disk, which is always satisfied.  
On the other hand, numerical simulations of gravito-turbulence
show that the second inequality is generally satisfied \citep{rice11}.  
Taken together, this means that in the outer regions of a realistic PPD,
cooling itself cannot cause instability.

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
We adopt the opacity scale $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$  as in
\citetalias{rafikov15}. We use 
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied
everywhere. Fig. \ref{rafikov_model} shows the equilirbium 
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation
(Eq. \ref{thindisk}).

Fig. \ref{rafikov_growth} show growth timescales and
optimum wavenumbers for viscous GI in this disk model.  
We also plot analytic estimates based on Eq. \ref{gammie_smallk} (instead of
Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) 
which gives the optimum wavenumber as
\begin{align}
  |K| = \frac{3}{2\theta Q},
\end{align}
with growth rate
\begin{align} 
  S = \frac{27\alpha}{16\theta^3Q^4}. 
\end{align}
These are similar to the isothermal results of
\citet[][their Eq. 19 and 21, respectively]{sterzik95}, and identical if one
takes $\theta=1$.

@@

\subsection{3D PPD with radiative diffusion}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?} 
We briefly consider 3D PPDs with explicit radiative diffusion
(\S\ref{rad_cool}). 
Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but without 
irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
wavenumber for $R\in[10,100]$AU; along with corresponding 2D 
results with softened self-gravity. There is a close match for
$R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
1$) and both models apply. However, beyond $60$AU where the disk
becomes optically-thin, radiative diffusion (the 3D curve) is not
valid and under-estimates the growth rates. Nevertheless, the
comparison shows that the transition radius $R_t\sim60$AU can be
calculated within the 2D framework. 
%This is not surprising since the
%instability is fundamentally driven by viscous and self-gravity forces in the
%momentum equations, not the energy equation.

>>>>

However, $\theta$ is now an effective 
irradiation level. 
%no longer a direct measure of
%irradiation. 
Eq. \ref{real_theta} shows that $\theta$ is related to the true
irradation temperature $\tirr$ through the function $g$ given by 
Eq. \ref{g_def}. This is
important because a physically non-irradiated disk with $\tirr=0$ will
still have a non-zero effective irradiation, $\theta\neq0$. 
%This results from the fact that PPD cooling generally depends on both
%$P$ and $\Sigma$, but in the absence of irradiation standard beta
%cooling only depends on $P$.

%On the
%other hand, fully irradiated disks with  $T=T_\mathrm{irr}$ have 
%$\theta=1$, as is in the previous problem formulation
%(Eq. \ref{tirr_def}).

@@

\subsection{Inviscid stability condition}
We note that for our adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
From the discussion in \S\ref{2d_inviscid} and applying
Eq. \ref{stable_condition},  we conclude that without viscous effects 
the disk is stable everywhere if  
\begin{align} 
  \gamma > \frac{3}{2} \quad \text{and} \quad Q >
  \sqrt{\frac{6}{5}} \label{ppd_invisc_cond} 
\end{align} 
are both satisfied.

PPDs become irradiation-dominated at large distances from the star,
where $T\to\tirr$, or $\theta\to 1$. {\bf ref?}
Then the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to
$\gamma > 1$ in the outer disk, which is always satisfied.  
On the other hand, numerical simulations of gravito-turbulence
show that $1\lesssim Q \lesssim 2$ \citep{gammie01,rice11}, and 
the second inequality is generally satisfied. Taken together, this
means that in the outer regions of a realistic PPD, cooling itself
cannot cause instability.

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
We adopt the opacity scale $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$  as in
\citetalias{rafikov15}. We use $\gamma=1.6$, approximately applicable
to molecular gas at low temperatures and used by other authors
\citep{rice11,baehr15}. Then the global inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied. 
%for outer disk choice of gamma doesn't matter 
%avoid complications from over-stable thermal modes 
Fig. \ref{rafikov_model} shows the equilirbium 
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation
(Eq. \ref{thindisk}).

Fig. \ref{rafikov_growth} show growth timescales and
optimum wavenumbers for viscous GI in this disk model.  
We also plot analytic estimates based on Eq. \ref{gammie_smallk} (instead of
Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) 
which gives the optimum wavenumber as
\begin{align}
  |K| = \frac{3}{2\theta Q},
\end{align}
with growth rate
\begin{align} 
  S = \frac{27\alpha}{16\theta^3Q^4}. 
\end{align}
These are similar to the isothermal results of
\citet[][their Eq. 19 and 21, respectively]{sterzik95}, and identical if one
takes $\theta=1$. %showing that irradiation makes the disk behave like
                  %isothermal disk

@@

\subsection{3D PPD with radiative diffusion}
%appropriate for not-so-large distances where irrad can be neglected 
We briefly consider 3D PPDs with explicit radiative diffusion
(\S\ref{rad_cool}). 
Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but 
with $T_\mathrm{irr}=0$, since our simple radiative diffusion model
does not include irradiation (\S\ref{rad_cool}). 
%physically, irrad is negligible for opt thick regions, and we are
%restricted to such regions if applying rad diff
We use a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
wavenumber for $R\in[10,100]$AU; along with corresponding 2D 
results with softened self-gravity. There is a close match for
$R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
1$) and both models apply. However, beyond $60$AU where the disk
becomes optically-thin, radiative diffusion (the 3D curve) is not
valid and under-estimates the growth rates. %; while the $f(\tau)$
%function approximately captures opticall-thin cooling in the 2D case.  
Nevertheless, the comparison shows that the transition radius
$R_t\sim60$AU can be calculated within the 2D framework.  
%This is not surprising since the
%instability is fundamentally driven by viscous and self-gravity forces in the
%momentum equations, not the energy equation.
